---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# DuckTinyCC

DuckTinyCC is a DuckDB C extension with a `tcc_*` SQL surface for TinyCC-based
in-process C scripting workflows.

## API Surface

All functionality is exposed through `tcc_module(...)` and selected with
`mode := ...`.  
Every mode returns one diagnostics row with `ok`, `code`, `message`, and
context columns.

### Session and Configuration

| Mode | Purpose | Notes |
|---|---|---|
| `config_get` | Show current session configuration | Includes runtime path and counters |
| `config_set` | Set session runtime path | Connection-scoped |
| `config_reset` | Reset runtime path and build state | Keeps extension loaded |
| `tcc_new_state` | Start a fresh build state | Drops staged build inputs |
| `list` | Show session registry/build counters | Diagnostics only |

### Build Inputs

| Mode | Purpose |
|---|---|
| `add_include`, `add_sysinclude` | Add include paths |
| `add_library_path`, `add_library` | Add library paths/libraries |
| `add_option`, `add_define` | Add compiler flags and defines |
| `add_header`, `add_source` | Stage C code units in session |
| `tinycc_bind` | Stage `symbol` + `sql_name` for next compile |

### Compile and Register

| Mode | Purpose | Notes |
|---|---|---|
| `compile`, `tinycc_compile` | Compile staged session code and register SQL UDF | Uses `tinycc_bind` or explicit `symbol`/`sql_name` |
| `quick_compile` | One-shot fast lane compile + register | Requires `source`, `symbol`, `sql_name`, `return_type`, `arg_types`; can also pass per-call `include_path`, `sysinclude_path`, `library_path`, `library`, `option`, `define_*`, `header` |

### Discovery Helpers

| Function | Purpose | Notes |
|---|---|---|
| `tcc_system_paths(runtime_path := '', library_path := '')` | Show include and library search paths used by TinyCC workflows | Includes existence flags; includes Windows/MSYS2/Rtools defaults on Windows builds |
| `tcc_library_probe(library := ..., runtime_path := '', library_path := '')` | Resolve a library short name or full filename/path | Supports `.a`, `.so*`, `.dylib`, `.dll`, `.lib` style names |

## Supported Types

Type metadata is explicit for `compile`, `tinycc_compile`, and
`quick_compile`:

- `return_type := ...` is required
- `arg_types := [...]` is required (use `[]` for zero arguments)

Current executable SQL registration support:

| Signature Piece | Accepted Tokens | DuckDB Type |
|---|---|---|
| `return_type` | `void`, `bool`, `i8`, `u8`, `i16`, `u16`, `i32`, `u32`, `i64`, `u64`, `f32`, `f64` | `BIGINT` for `void`; otherwise matching scalar type |
| each `arg_types` element | `bool`, `i8`, `u8`, `i16`, `u16`, `i32`, `u32`, `i64`, `u64`, `f32`, `f64` | matching scalar type |

Additional limits and behavior:

- Arity: `0..10` arguments
- NULL handling: NULL-in/NULL-out
- `void` return registers as a SQL function that emits `NULL` values

## Build

```sh
make configure
make debug
```

## Test

```sh
make test_debug
make test_release
```

## Examples

The examples below use `DBI` + `duckdb` and focus on one workflow at a time.

### 1) Setup

This chunk connects to DuckDB in-memory and loads the extension artifact.

```{r eval=TRUE}
library(DBI)
library(duckdb)

drv <- duckdb::duckdb(config = list(allow_unsigned_extensions = "true"))
con <- dbConnect(drv, dbdir = ":memory:")
ext_path <- normalizePath("build/release/ducktinycc.duckdb_extension", mustWork = FALSE)
dbExecute(con, sprintf("LOAD '%s'", ext_path))
```

### 2) Session Configuration

This chunk initializes a clean session and inspects the current state.

```{r eval=TRUE}
dbGetQuery(con, "
  SELECT ok, mode, code, detail
  FROM tcc_module(
    mode := 'config_set',
    runtime_path := ''
  )
")

dbGetQuery(con, "
  SELECT ok, mode, code, detail
  FROM tcc_module(mode := 'config_get')
")

dbGetQuery(con, "
  SELECT ok, mode, code, detail
  FROM tcc_module(mode := 'list')
")
```

### 3) System Paths and Library Probe Helpers

This chunk inspects computed search paths and probes a static library by name.

```{r eval=TRUE}
dbGetQuery(con, "
  SELECT kind, key, exists, value
  FROM tcc_system_paths()
  LIMIT 12
")

dbGetQuery(con, "
  SELECT kind, key, exists, value, detail
  FROM tcc_library_probe(library := 'libtcc1.a')
")
```

### 4) Example A: Staged Build + `tinycc_bind` + `tinycc_compile`

This example stages shared build inputs once, compiles two symbols, and calls both SQL functions.

```{r eval=TRUE}
dbGetQuery(con, "SELECT ok, mode, code, detail FROM tcc_module(mode := 'tcc_new_state')")
dbGetQuery(con, "SELECT ok, mode, code, detail FROM tcc_module(mode := 'add_include', include_path := 'third_party/tinycc/include')")
dbGetQuery(con, "SELECT ok, mode, code, detail FROM tcc_module(mode := 'add_sysinclude', sysinclude_path := 'third_party/tinycc/include')")
dbGetQuery(con, "SELECT ok, mode, code, detail FROM tcc_module(mode := 'add_library_path', library_path := 'third_party/tinycc')")
dbGetQuery(con, "SELECT ok, mode, code, detail FROM tcc_module(mode := 'add_option', option := '-O2')")
dbGetQuery(con, "SELECT ok, mode, code, detail FROM tcc_module(mode := 'add_define', define_name := 'TCC_SHIFT', define_value := '3')")

dbGetQuery(con, "
  SELECT ok, mode, code, detail
  FROM tcc_module(
    mode := 'add_header',
    header := '#include <stdint.h>'
  )
")

dbGetQuery(con, "
  SELECT ok, mode, code, detail
  FROM tcc_module(
    mode := 'add_source',
    source := '
      long long tcc_add_shift(long long x){ return x + TCC_SHIFT; }
      long long tcc_times2(long long x){ return x * 2; }
    '
  )
")

dbGetQuery(con, "
  SELECT ok, mode, code
  FROM tcc_module(mode := 'tinycc_bind', symbol := 'tcc_add_shift', sql_name := 'tcc_add_shift')
")
dbGetQuery(con, "
  SELECT ok, mode, code
  FROM tcc_module(
    mode := 'tinycc_compile',
    return_type := 'i64',
    arg_types := ['i64']
  )
")
dbGetQuery(con, "SELECT tcc_add_shift(39) AS value")

dbGetQuery(con, "
  SELECT ok, mode, code
  FROM tcc_module(mode := 'tinycc_bind', symbol := 'tcc_times2', sql_name := 'tcc_times2')
")
dbGetQuery(con, "
  SELECT ok, mode, code
  FROM tcc_module(
    mode := 'compile',
    return_type := 'i64',
    arg_types := ['i64']
  )
")
dbGetQuery(con, "SELECT tcc_times2(21) AS value")
```

### 5) Example B: Fast Lane `quick_compile` (with include/library inputs)

This example compiles and registers directly from one SQL call, including
`include_path`, `library_path`, and `library`.

```{r eval=TRUE}
dbGetQuery(con, "
  SELECT ok, mode, code
  FROM tcc_module(
    mode := 'quick_compile',
    source := '#include <math.h>
double qpow(double x, double y){ return pow(x, y); }',
    symbol := 'qpow',
    sql_name := 'qpow',
    return_type := 'f64',
    arg_types := ['f64', 'f64'],
    include_path := 'third_party/tinycc/include',
    library_path := 'third_party/tinycc',
    library := 'm'
  )
")
dbGetQuery(con, "SELECT CAST(qpow(2.0, 5.0) AS BIGINT) AS value")
```

### 6) Example C: Libraries (`add_library`)

This example links `libm`, compiles a function that uses `pow`, then calls it.
The required C header is included in the source unit.

```{r eval=TRUE}
dbGetQuery(con, "SELECT ok, mode, code FROM tcc_module(mode := 'tcc_new_state')")
dbGetQuery(con, "SELECT ok, mode, code FROM tcc_module(mode := 'add_library', library := 'm')")
dbGetQuery(con, "
  SELECT ok, mode, code
  FROM tcc_module(
    mode := 'add_source',
    source := '#include <math.h>
double pwr(double x, double y){ return pow(x, y); }'
  )
")
dbGetQuery(con, "
  SELECT ok, mode, code
  FROM tcc_module(mode := 'tinycc_bind', symbol := 'pwr', sql_name := 'pwr')
")
dbGetQuery(con, "
  SELECT ok, mode, code
  FROM tcc_module(mode := 'compile', return_type := 'f64', arg_types := ['f64', 'f64'])
")
dbGetQuery(con, "SELECT CAST(pwr(2.0, 5.0) AS BIGINT) AS value")
```

### 7) Reset Session

This chunk resets session state and shows summary counters.

```{r eval=TRUE}
dbGetQuery(con, "SELECT ok, mode, code, detail FROM tcc_module(mode := 'list')")
dbGetQuery(con, "SELECT ok, mode, code FROM tcc_module(mode := 'config_reset')")
```

### 8) CLI Benchmark Snippet

Use the DuckDB CLI timer to benchmark compile and call latency quickly:

```{bash eval=TRUE, collapse=FALSE, comment=""}
duckdb -unsigned <<'SQL'
.timer on
LOAD 'build/release/ducktinycc.duckdb_extension';
SELECT ok, mode, code
FROM tcc_module(
  mode := 'quick_compile',
  source := 'int add_i32(int a, int b){ return a + b; }',
  symbol := 'add_i32',
  sql_name := 'add_i32',
  return_type := 'i32',
  arg_types := ['i32', 'i32']
);
SELECT add_i32(20, 22);
SELECT SUM(add_i32(i::INTEGER, 42::INTEGER)) AS s FROM range(1000000) t(i);
SELECT SUM(add_i32(i::INTEGER, 42::INTEGER)) AS s FROM range(1000000) t(i);
SELECT SUM(add_i32(i::INTEGER, 42::INTEGER)) AS s FROM range(1000000) t(i);
SQL
```

### 9) Cleanup

This chunk closes the database connection.

```{r eval=TRUE}
dbDisconnect(con, shutdown = TRUE)
```

## Notes

- Function invocation is plain SQL (`SELECT my_udf(...)`) after `compile` or `quick_compile`.
- `add_header`/`add_source` are compiled as separate translation units. For external library prototypes, include headers directly in the relevant `source` unit.
- Use `tcc_system_paths()` and `tcc_library_probe()` to diagnose platform-specific library resolution (including Windows/Rtools/MSYS2 layouts).
- All outputs are returned as a diagnostics table for SQL-native observability.
